Categories of Optics

### 2018-06-08 paper.tex
Start reworking examples

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fmatching}{\textsc{Matching}}

@@

%\todo{Past work:
%\begin{itemize}
%\item The Haskell lens people: edwardk, shachaf, xplat, roconnor... (TODO: find out who these people are)
%\item The Doubles paper
%\item Bartosz's blog post which was written about the same time I started this. (Only tensor-like optics, no discussion of laws, confusion about enrichment(?))
%\item The profunctor optics paper out of Oxford. (No generality, nothing categorical, no discussion of laws)
%\end{itemize}
%}
%
%\todo{Contributions:
%\begin{itemize}
%\item ...
%\end{itemize}
%}
%
%\todo{Introduction should include:
%\begin{itemize}
%\item Previous attempts to generalise optics
%\item Some remark about how it's hard to isolate exactly what $\Set$-like properties are being used
%\end{itemize}
%}
A lens from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$ that typically obeys three laws (whose names are in diagrammatic order):
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be represented faithfully in $S$;
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times \id_A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}

@@

\begin{proposition}
A monoidal functor $F : \C \to \D$ induces a  functor $\Optic(F) : \Optic_\C \to \Optic_\D$
\end{proposition}
\begin{proof}
On objects this is simply $\Optic(F)(S, S') = (FS, FS')$. On a morphism $\langle l, r \rangle : (S, S') \to (A, A')$, we define
\begin{align*}
F\langle l, r \rangle := \langle \phi^{-1}_{M,A} (Fl), (Fr) \phi_{M,A'}\rangle
\end{align*}
where $\phi_{M,A} : FM \otimes FA \to F(M \otimes A)$ is the structure map of the monoidal functor.
\todo{functoriality}
\todo{monoidality}
\end{proof}

@@

\begin{definition}
A \emph{teleological functor} $F : \C \to \D$ is a strong symmetric monoidal functor that restricts to a functor $F_d : \C_d \to \D_d$ on the dualisable subcategories, and such that $F(\varepsilon_X) = \varepsilon_{FX}$ \todo{up to the structure isomorphisms of $F$...}
\end{definition}

@@

@@

\section{Lawful Optics}
\todo{intro}

\begin{remark}
In this section we will focus only on optics of the form $p : (S,S) \hto (A, A)$. We therefore abbreviate $\Optic_\M((S, S), (A, A))$ as $\Optic_\M(S, A)$ and $p : (S, S) \hto (A, A)$ as $p : S \hto A$.
\end{remark}

For a fixed $M \in \M$, there is a map $\C(S, M A) \times \C(M A, S) \to \C(S, S)$ given by composition. This induces a map $outside : \Optic_\M(S, A) \to \C(S, S)$. By composition in the other direction we have a map $\C(S, M A) \times \C(M A, S) \to \C(M A, M A)$. Including this object $\C(M A, M A)$ into the coend $\int^{M \in \M} \C(M A, M A)$, we induce a map $inside : \Optic_\M(S, A) \to \int^{M \in \M} \C(M A, M A)$. 
\begin{definition}
An optic $p : S \hto A$ is \emph{lawful} if $outside(p) = \id_S$, and $inside(p) = \langle \phi A \rangle$ for some $\phi : M \to M$ in $\M$.
\end{definition}

\begin{proposition}
There is an subcategory $\Lawful_\M$ of $\Optic_\M$ given by objects of the form $(X, X)$ and lawful optics between them.
\end{proposition}
\begin{proof}
The identity optic $\iota(\id_X, \id_X)$ is clearly lawful.

Suppose we have two lawful optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$, where $l : R \to MS$ and $l' : S \to NA$, and such that $lr = \phi S$ and $l'r' = \psi A$. Then the composite $\langle (Ml')l, r (Mr')  \rangle$ is also a lawful optic:
\begin{align*}
r (Mr')(Ml')l &= r M(r'l') l = rl = \id_S
\intertext{and}
(Ml')l r (Mr') &= (Ml')(\phi S)(Mr') \\ 
&= (\phi NA) M(l')M(r') \\ 
&= (\phi NA) M(l'r') \\
&= (\phi NA) M(\psi A) \\
&= \phi(\psi A)
\end{align*}
by the naturality of the action of $\M$.
\end{proof}

\begin{proposition}
\label{prop-onthenose}
If $p : S \to A$ is a lawful optic, then there exists a representative $p = \langle l, r \rangle$ such that $rl = \phi A$ on the nose for some $\phi : M \to M$.
\end{proposition}
\begin{proof}
The coend relation in $\int^{M \in \M} \C(M A, M A)$ states that two morphisms $f : M A \to M A$ and $g : N A \to N A$ are identified when there exists a $k : N A \to M A$ and morphism $\phi : M \to N$ such that
\[
\begin{tikzcd}
M A \ar[r, "\phi_A"] \ar[d, "f", swap] & N A \ar[d, "g"] \ar[dl, "k"] \\
M A \ar[r, "\phi_A", swap] & N A
\end{tikzcd}
\]
commutes. This gives a binary relation $f \rightsquigarrow g$ that is not likely to be symmetric or transitive in general. The coend is obtained by identifying all morphisms related in this way, $\langle f \rangle = \langle g \rangle$ iff there exists a zigzag $f = u_1 \leftrightsquigarrow u_2 \leftrightsquigarrow \dots  \leftrightsquigarrow u_{n-1} \leftrightsquigarrow u_n = g$ where the relation may apply in either direction.

@@

In view of this Proposition, whenever we posit a representative $\langle l, r \rangle$ of an lawful optic, we may assume that $lr = \phi A$ on the nose, for some $\phi : M \to M$.

@@

The general pattern is, once we choose a particular monoidal action $\M \to [\C, \C]$, we can find an isomorphism between the set of optics $(S, S') \hto (A, A')$ and a set that is easier to describe. The goal is then to find necessary and sufficient conditions on this second set for the corresponding element of the coend to be a lawful optic.

@@

The optics for this action are exactly pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\Optic_\times((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \times A) \times \C(M \times A', S') \\
&\cong \int^{M \in \C} \C(S, M) \times \C(S, A) \times \C(M \times A', S') && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A', S') && \text{(co-Yoneda)}
\end{align*}

Explicitly this isomorphism states that, given an (unlawful) lens $\langle l, r \rangle : (S, S') \hto (A, A')$, the corresponding $\fget$ and $\fput$ functions are given by $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. In the other direction, given $\fget$ and $\fput$, the corresponding element in the coend has representative $\langle l, r \rangle$, where $l = [\id_s, \fget]$ and $r = \fput$. Note that even if the $\fget$ and $\fput$ correspond to a law-abiding lens, this particular representative will not be an isomorphism.

@@

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
Suppose $\langle l, r \rangle : S \hto A$ is a lawful lens in our sense, i.e. that $rl = \id_S$ and $\langle lr \rangle = \langle \phi \times A \rangle$ for some $\phi : M \to M$. Then $\fput$ and $\fget$ for this lens obey the three lens laws.
\end{proposition}
\begin{proof}
By Proposition \ref{prop-onthenose}, we may assume that $lr = \phi \times A$.

Let $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. Verification of the laws is straightforward:
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (t\pi_1 l \times A) \times A) \\
&= r (t\pi_1 l \times A) \pi_{1,3} \\
&= r lr(\pi_1 l \times A) \pi_{1,3} \\
&= r(\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}
\end{proof}

This is well known in the case $lr$ is the identity \cite{IsomorphismLensesPost}, we note here that the weaker condition is sufficient.

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a lawful lens.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has as a representative $\langle [\id_s, \fget], \fput \rangle$.

Let $\fput_x : S \to S$ be the composite $S \to S \times 1 \xrightarrow{x} S \times A \xrightarrow{\fput} S$. The commutative diagram
\[
\begin{tikzcd}
S \times A \ar[r, "\fput_x \times A "] \ar[d, "lr", swap] & S \times A \ar[d, "\fput_x \times A"] \ar[dl, "lr"] \\
S \times A \ar[r, "\fput_x \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle lr \rangle = \langle \fput_x \times A \rangle$ in $\int^{M \in \M} \C(M A, M A)$.
\todo{Is there some property of $\times$ used here we can abstract? Having a map $1 \to A$ guarantees that $- \times A$ is faithful, is this the trick?}
\end{proof}

@@

\subsection{Non-Cartesian Lenses}
If $\C$ is closed monoidal, we can use a similar trick to eliminate the coend in the definition of an optic:

\begin{align*}
\Optic_\otimes((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \\
&\cong \int^{M \in \C} \C(S, M \otimes A) \times \C(M, \homC(A',S')) \\
&\cong \C(S, \homC(A',S') \otimes A)
\end{align*}
where $\homC(A', S')$ denotes the internal hom. If $\C$ is cartesian, this is of course isomorphic to the pair of maps discussed in the previous section.

\todo{laws?}
\todo{The use of lenses in a type system with linear types seems worth exploring!}

@@

Suppose $\C$ is cartesian closed and has finite coproducts. Then $\C$ acts on itself via $M \cdot A = (M \sqcup 1) \times A$.
\begin{align*}
\Optic((S, S'), (A, A')) 
&= \int^{M \in \C} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
&= \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
&= \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S') \\
\end{align*}

\todo{Equivalently(?), let $\M$ be the category of pointed sets}

@@

To define setters we must first recall the notion of a strong functor.

>>>>

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fmatching}{\textsc{Matching}}
\newcommand{\funzip}{\textsc{Unzip}}

\newcommand{\inside}{\mathsf{inside}}
\newcommand{\outside}{\mathsf{outside}}
\newcommand{\once}{\mathsf{once}}
\newcommand{\twice}{\mathsf{twice}}

@@

\todo{Past work:
\begin{itemize}
\item The Haskell lens people: edwardk, shachaf, xplat, roconnor... (TODO: find out who these people are)
\item The Doubles paper
\item Bartosz's blog post which was written about the same time I started this. (Only tensor-like optics, no discussion of laws, confusion about enrichment(?))
\item The profunctor optics paper out of Oxford. (No generality, nothing categorical, no discussion of laws)
\item Very old Dialectica category stuff
\end{itemize}
}

\todo{Contributions:
\begin{itemize}
\item A definition of optics general enough to encompass lenses, setters, traversals, and others?
\item A universal property satisfied by the category of optics in the monoidal case
\item A satisfying categorical description of lens laws
\item An attempt to isolate exactly what $\Set$-like properties are being used in various places
\item A couple of new avenues for generalising optics
\end{itemize}
}

A lens from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$ that typically obeys three laws (whose names are in diagrammatic order):
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be represented faithfully in $S$;
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}

@@

\begin{proposition}
\label{prop-change-of-action-monoidal}
A monoidal functor $F : \C \to \D$ induces a functor $\Optic(F) : \Optic_\C \to \Optic_\D$.
\end{proposition}
\begin{proof}
On objects this is simply $\Optic(F)(S, S') = (FS, FS')$. On a morphism $\langle l, r \rangle : (S, S') \to (A, A')$, we define
\begin{align*}
F\langle l, r \rangle := \langle \phi^{-1}_{M,A} (Fl), (Fr) \phi_{M,A'}\rangle
\end{align*}
where $\phi_{M,A} : FM \otimes FA \to F(M \otimes A)$ is the structure map of the monoidal functor.
\todo{functoriality}
\todo{monoidality}
\end{proof}

@@

\begin{definition}
A \emph{teleological functor} $F : \C \to \D$ is a symmetric monoidal functor that restricts to a functor $F_d : \C_d \to \D_d$ on the dualisable subcategories, and such that $F(\varepsilon_X) = \varepsilon_{FX}$ \todo{up to the structure isomorphisms of $F$...}
\end{definition}

@@

\begin{proposition}
\label{prop-change-of-action}
Suppose $\C$ and $\D$ are categories acted on by $\M$ and $\mathcal{N}$ respectively. If $F : \C \to \D$ is a functor that commutes with the actions, then there is an induced functor $\Optic(F) : \Optic_\M \to \Optic_\N$.
\end{proposition}
\begin{proof}
Analogous to Proposition \ref{prop-change-of-action-monoidal}
\end{proof}

@@

\section{Lawful Optics}
\todo{intro}
\todo{Here I am using ; to mean composition in diagrammatic order, I need to make a final decision on which to use}

\begin{remark}
In this section we will focus only on optics of the form $p : (S,S) \hto (A, A)$. For brevity we will write $\Optic_\M((S, S), (A, A))$ as $\Optic_\M(S, A)$ and $p : (S, S) \hto (A, A)$ as $p : S \hto A$.
\end{remark}

\todo{In this section I am using semicolon to mean composition in diagrammatic order. Should I also switch the comma in $\langle l, r \rangle$ to be a $\mid$?}

For a fixed $M \in \M$, there is a map $\C(S, M A) \times \C(M A, S) \to \C(S, S)$ given by composition. This induces a map $\outside : \Optic_\M(S, A) \to \C(S, S)$, given equationally by $\outside(\langle l, r \rangle) = l ; r$

There are also maps two maps 
\begin{align*}
\once, \twice : \Optic_\M(S, A) \to \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)
\end{align*}
defined by 
\begin{align*}
\once(\langle l, r \rangle) &= \langle l, \id_{MA}, r \rangle \\
\twice(\langle l, r \rangle) &= \langle l, r;l, r \rangle
\end{align*}

\begin{definition}
An optic $p : S \hto A$ is \emph{lawful} if 
\begin{align*}
\outside(p) &= \id_S \\
\once(p) &= \twice(p)
\end{align*}
\end{definition}

We will see that in all the standard examples of optics, this specialises to the expected laws for that type of optic.

\begin{proposition}
There is an subcategory $\Lawful_\M$ of $\Optic_\M$ given by objects of the form $(S, S)$ and lawful optics between them.
\end{proposition}
\begin{proof}
The identity optic $\iota(\id_S, \id_S)$ is lawful: 
\begin{align*}
\outside(\langle \lambda^{-1}_S, \lambda_{S} \rangle) &= \lambda_{S} \lambda^{-1}_S = \id_S \\
\intertext{and}
\once(\langle \lambda^{-1}_S, \lambda_{S} \rangle) 
&= \langle \lambda^{-1}_S, \id_{IA}, \lambda_{S} \rangle \\
&= \langle \lambda^{-1}_S, \lambda_{S};\lambda^{-1}_S , \lambda_{S} \rangle \\
&= \twice(\langle \lambda^{-1}_S, \lambda_{S} \rangle)
\end{align*}

Suppose we have two lawful optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$, where $l : R \to MS$ and $l' : S \to NA$. We must show that $\langle (Ml')l, r (Mr')  \rangle$ is also a lawful optic. The first equation is straightforward:
\begin{align*}
\outside(\langle l; (Ml'),(Mr') ; r   \rangle) 
&= l ; Ml' ; Mr' ; r \\
&= l ; M(l'r') ; r \\
&= l ; M\id_{NA} ; r \\
&= l ; r \\
&= \id_{MNA}
\end{align*}

For the second, we must show
\[ \langle l;(Ml'), (Mr'); r;l;(Ml') , (Mr') ; r  \rangle = \langle l;(Ml'), \id_{MNA} , (Mr') ; r  \rangle. \]

We do this by showing that, in some sense, the expressions in the above equation respect the coend relations defining the codomains of $\once(\langle l, r \rangle)$ and $\once(\langle l', r' \rangle)$.

So consider the generating relation $\langle l;\phi_S, c, r \rangle = \langle l, \phi_S ; c, r \rangle$ in \[\int^{M_1, M_2 \in \M} \C(R, M_1 S) \times \C(M_1 S, M_2 S) \times \C(M_2 S, R).\]
This implies that 
\begin{align*}
\langle l;\phi_S;(Ml'), (Mr'); c ;(Ml') , (Mr') ; r \rangle 
&= \langle l;(Ml');\phi_{NA}, (Mr'); c ;(Ml') , (Mr') ; r \rangle && \text{(functoriality)} \\
&= \langle l;(Ml'), \phi_{NA};(Mr'); c ;(Ml') , (Mr') ; r \rangle && \text{(coend relation)} \\
&= \langle l;(Ml'), (Mr');\phi_{S}; c ;(Ml') , (Mr') ; r \rangle && \text{(functoriality)}
\end{align*}
And similarly for the other generating relation, $\langle l, c;\phi_S, r \rangle = \langle l,  c, \phi_S; r \rangle$.

Therefore, the chain of relations proving $\langle l, r;l, r \rangle = \langle l, \id_{MA}, r \rangle$ can be replicated to show
\begin{align*}
\langle l;(Ml'), (Mr');r;l;(Ml') , (Mr') ; r \rangle &= \langle l;(Ml'), (Mr');\id_{MA};(Ml') , (Mr') ; r \rangle,
\end{align*}
which is equal to $\langle l;(Ml'), M(r'l'), (Mr') ; r \rangle$.

Similarly, a generating relation $\langle l';\psi_A, c', r' \rangle = \langle l', \psi_A ; c', r' \rangle$ in \[\int^{N_1, N_2 \in \M} \C(S, N_1 A) \times \C(N_1 A, N_2 A) \times \C(M_2 A, S)\]
implies that
\begin{align*}
\langle l;M(l';\psi_A), Mc', (Mr') ; r \rangle 
&= \langle l;Ml';M\psi_A, Mc', (Mr') ; r \rangle \\
&= \langle l;Ml', M\psi_A ; Mc', (Mr') ; r \rangle \\
&= \langle l;Ml', M(\psi_A;c'), (Mr') ; r \rangle 
\end{align*}
So again we can replicate the chain of relations proving $\langle l', r';l', r' \rangle = \langle l', \id_{MA}, r' \rangle$ to show that
\[\langle l;(Ml'), M(r'l'), (Mr') ; r \rangle = \langle l;(Ml'), \id_{MNA}, (Mr') ; r \rangle \]

We conclude that composition preserves lawfulness, so $\Lawful_\M$ is indeed a subcategory of $\Optic_\M$.

\todo{There is probably a more conceptual way to do this}
\end{proof}

The requirement that $\once(p) = \twice(p)$ is mysterious, but there are sufficient conditions that are easier to verify.

\begin{proposition}
Let $\langle l, r \rangle : S \hto A$ be an optic. If $l;r = \id_S$ and $r;l = \phi A$ for some $\phi : M \to M$ in $\M$, then $\langle l, r \rangle$ is a lawful optic.
\end{proposition}
\begin{proof}
The $\outside(\langle l, r \rangle) = l;r = \id_S$ is exactly the first condition. And for the second, we verify:
\begin{align*}
\twice(\langle l, r \rangle) 
&= \langle l, r;l, r \rangle \\
&= \langle l, \phi A, r \rangle && \text{($r;l = \phi A$)}\\
&= \langle l (\phi A), \id_{MA}, r \rangle && \text{(coend relation)} \\
&= \langle l;r;l, \id_{MA}, r \rangle && \text{($r;l = \phi A$ again)}\\
&= \langle l, \id_{MA}, r \rangle  && \text{($l;r = \id_S$)}\\
&= \once(\langle l, r \rangle) 
\end{align*}
\end{proof}

\begin{remark}
Even if $r;l = \phi A$ for some $\phi$, the same is not necessarily true of any other representative of the same optic.
\end{remark}

Let $\inside : \Optic_\M(S, A) \to \int^{M \in \M} \C(M A, M A)$ be the map induced by $\inside(\langle l, r \rangle) = \langle r ; l \rangle$. We might ask that instead of requiring $r;l = \phi A$ exactly, we have $\langle r ; l \rangle = \langle \phi A \rangle$ in $\int^{M \in \M} \C(M A, M A)$2  . In fact, this is equivalent:

\begin{proposition}
\label{prop-onthenose}
Suppose $p : S \hto A$ satisfies $\outside(p) = \id_S$ and $\inside(p) = \langle \phi A \rangle$. Then there exists a representative $\langle l, r \rangle$ such that $l ; r = \psi A$ on the nose for some (possibly different) $\psi : M \to M$.
\end{proposition}
\begin{proof}
\todo{Go through this proof for formatting again.}

The coend relation in $\int^{M \in \M} \C(M A, M A)$ states that two morphisms $f : M A \to M A$ and $g : N A \to N A$ are identified when there exists a $k : N A \to M A$ and morphism $\phi : M \to N$ such that
\[
\begin{tikzcd}
M A \ar[r, "\phi_A"] \ar[d, "f", swap] & N A \ar[d, "g"] \ar[dl, "k"] \\
M A \ar[r, "\phi_A", swap] & N A
\end{tikzcd}
\]
commutes. This gives a binary relation $f \rightsquigarrow g$ that is not likely to be symmetric or transitive in general. The coend is obtained by identifying all morphisms related in this way, $\langle f \rangle = \langle g \rangle$ iff there exists a zigzag $f = u_1 \leftrightsquigarrow u_2 \leftrightsquigarrow \dots  \leftrightsquigarrow u_{n-1} \leftrightsquigarrow u_n = g$ where the relation may apply in either direction.

@@

Finally, lawfulness is preserved by the change-of-action functors of Proposition \ref{prop-change-of-action}.

\begin{proposition}
If $F : \C \to \D$ is a functor that commutes with the monoidal actions on $\C$ and $\D$, then $\Optic(F) : \Optic_\M \to \Optic_\N$ restricts to a functor $\Lawful_\M \to \Lawful_\N$.
\end{proposition}
\begin{proof}
\todo{this should be in here}
\end{proof}

@@

The general pattern is as follows. Once we choose a particular monoidal action $\M \to [\C, \C]$, we find an isomorphism between the set of optics $(S, S') \hto (A, A')$ and a set $\mathbb{O}((S, S'), (A, A'))$ that is easier to describe. Ideally, we then find a similar description $\mathbb{D}(S, A)$ for the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)\] appearing in the definition of a lawful optic. We can then ``read off'' what conditions on an element of $\mathbb{O}((S, S), (A, A))$ are needed to ensure that the corresponding element of $\Optic_\M(S, A)$ is lawful.

@@

The optics for this action correspond exactly to pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\Optic_\times((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \times A) \times \C(M \times A', S') \\
&\cong \int^{M \in \C} \C(S, M) \times \C(S, A) \times \C(M \times A', S') && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A', S') && \text{(co-Yoneda)} \\
&=: \mathbb{O}((S, S'), (A, A'))
\end{align*}

Explicitly this isomorphism states that, given an optic $\langle l, r \rangle : (S, S') \hto (A, A')$, the corresponding element of $\mathbb{O}((S, S'), (A, A'))$ is the pair $(\fget, \fput)$, where $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. In the other direction, given $(\fget, \fput)$, the corresponding optic has representative $\langle [\id_S, \fget],  \fput \rangle$.

@@

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
A lens $p : S \hto A$ is a lawful lens in our sense, i.e. that $\outside(p) = \id_S$ and $\once(p) = \twice(p)$, if and only if the corresponding $\fput$ and $\fget$ for this lens obey the three lens laws.
\end{proposition}
\begin{proof}
We begin by giving
\[ \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S)\]
the same treatment as $\Optic_\times(S, A)$. By the same sorts of arguments as above \todo{is this worth writing out?}, this is isomorphic to
\[\mathbb{D}(S, A) := \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)\]
with the isomorphism given by 
\begin{align*}
\Phi(\langle l, c, r \rangle) = (&\pi_2 l, \\
&\pi_2 c (\pi_1 l \times A), \\
&r(\pi_1 c (\pi_1 l \times A) \times A))
\end{align*}
Now suppose we are given a lens $(\fget, \fput)$, which in coend form is $\langle [\id_S, \fget],  \fput \rangle$. Evaluating $\outside$ on this gives:
\begin{align*}
\outside(\langle [\id_S, \fget],  \fput \rangle) = \fput [\id_S, \fget]
\end{align*}
so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

We now have to slog through evaluating $\Phi(\once(p))$ and $\Phi(\twice(p))$.
\begin{alignat*}{3}
\Phi(\once(\langle [\id_S, \fget],  \fput \rangle)) &= 
\Phi(&&\langle [\id_S, \fget], \id_{S \times A}, \fput \rangle) \\
&= (&&\pi_2 [\id_S, \fget], \\
&&&\pi_2 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A), \\
&&& \fput (\pi_1 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A) \times A) \quad) \\
%
&= (&&\fget, \\
&&& \pi_2, \\
&&& \fput \pi_{1,3} \quad) \\
%%%%
%%%%
\Phi(\twice(\langle [\id_S, \fget],  \fput \rangle)) &= 
\Phi(&&\langle [\id_S, \fget], [\id_S, \fget]\fput, \fput \rangle) \\
&= (&&\pi_2 [\id_S, \fget], \\
&&&\pi_2 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A), \\
&&& \fput (\pi_1 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A) \times A)) \quad) \\
%
&= (&&\fget, \\
&&& \fget \; \fput (\id_S \times A), \\
&&& \fput (\fput (\id_S \times A) \times A) \quad) \\
%
&= (&&\fget, \\
&&& \fget \; \fput, \\
&&& \fput (\fput \times A) \quad)
\end{alignat*}
So $\Phi(\once(p))$ being equal to $\Phi(\twice(p))$ is exactly equivalent to the $\fput\fget$ and $\fput\fput$ laws holding.
\end{proof}

\todo{This last proposition should probably be deleted now}

@@

\subsection{Linear Lenses}
\newcommand{\ev}{\mathsf{ev}}
\newcommand{\coev}{\mathsf{coev}}

If $\C$ is closed monoidal, we can use a similar trick to eliminate the coends.

\begin{align*}
\Optic_\otimes((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \\
&\cong \int^{M \in \C} \C(S, M \otimes A) \times \C(M, \homC(A',S')) \\
&\cong \C(S, \homC(A',S') \otimes A) \\
&=: \mathbb{O}((S, S'), (A, A'))
\end{align*}
where $\homC(A', S')$ denotes the internal hom. If $\C$ is cartesian, this is of course isomorphic to the pair of maps discussed in the previous section.

For an optic, the corresponding element of $\mathbb{O}((S, S'), (A, A'))$ will be called $\funzip$. The above isomorphism sends $\funzip$ to the element $\langle \funzip , \ev_{A', S'} \rangle$, where $\ev_{A', S'} : \homC(A',S') \otimes A' \to S'$ is the evaluation map. In the other direction, given $\langle l, r \rangle$ we have $\funzip = (\tilde r \otimes A)l$ where $\tilde r : M \to \homC(A', S')$ denotes the transpose of $r$.

We cannot possibly use the three $\fput$/$\fget$ style lens laws in this setting as we lack projections, but using the same techniques as for cartesian lenses above we can devise some new ones:

\begin{proposition}
A linear lens $p : S \hto A$ is lawful iff the following two laws for $\funzip$ hold:
\begin{align*}
\ev_{A, S} \; \funzip &= \id_S && \textsc{(Rezip)} \\
(\coev_{\homC(A, S), A} \otimes A)\funzip &= ((\funzip \circ -) \otimes A)\funzip && \textsc{(ZipZip)} 
\end{align*}
where \[ \funzip \circ - : \homC(A, S) \to \homC(A, \homC(A, S) \otimes A) \] denotes the internal composition. \todo{I obviously need better names for the laws}
\end{proposition}
\begin{proof}
We have an isomorphism:
\begin{align*}
& \int^{M_1, M_2 \in \C} \C(S, M_1 \otimes A) \times \C(M_1 \otimes A, M_2 \otimes A) \times \C(M_2 \otimes A, S) \\
&\cong \int^{M_1, M_2 \in \C} \C(S, M_1 \otimes A) \times \C(M_1, \homC(A, M_2 \otimes A)) \times \C(M_2, \homC(A, S)) \\
&\cong \int^{M_1 \in \C} \C(S, M_1 \otimes A) \times \C(M_1, \homC(A, \homC(A, S) \otimes A)) \\
&\cong \C(S, \homC(A, \homC(A, S) \otimes A) \otimes A)
\end{align*}
which evaluated on an element $\langle l, c, r \rangle$ is given by
\begin{align*}
\Phi(\langle l, c, r \rangle) &= (\widetilde{(\tilde{r} \otimes A)c} \otimes A)l
\end{align*}
Now suppose we are given an optic $p$ with corresponding element of $\mathbb{O}(S, A)$ is $\funzip$. Then $\outside(\langle \funzip , \ev_{A, S} \rangle) = \ev_{A, S} \; \funzip$, giving the first law.

Now we evaluate $\Phi(\once(p))$ and $\Phi(\twice(p))$.
\begin{align*}
\Phi(\once(\langle \funzip , \ev_{A, S} \rangle)) 
&= \Phi(\langle \funzip , \id_{\homC(A,S) \otimes A}, \ev_{A, S} \rangle) \\
&= (\widetilde{(\tilde{\ev_{A, S}} \otimes A)\id_{\homC(A,S) \otimes A}} \otimes A)\funzip \\
&= (\coev_{\homC(A, S), A} \otimes A)\funzip \\
\Phi(\twice(\langle \funzip , \ev_{A, S} \rangle)) 
&= \Phi(\langle \funzip , \funzip \; \ev_{A, S}, \ev_{A, S} \rangle) \\
&= (\widetilde{(\tilde{\ev_{A, S}} \otimes A)(\funzip \; \ev_{A, S})} \otimes A)\funzip \\
&= (\widetilde{(\funzip \; \ev_{A, S})} \otimes A)\funzip \\
&= ((\funzip \circ -) \otimes A)\funzip
\end{align*}
\todo{Fix those overtildes}
So indeed the optic is lawful iff $(\coev_{\homC(A, S), A} \otimes A)\funzip = ((\funzip \circ -) \otimes A)\funzip$.
\end{proof}

\todo{The use of lenses in a language with linear types seems worth exploring!}

@@

Suppose $\C$ is cartesian closed and has finite coproducts. Then $\C$ acts on itself via $M \cdot A = (M \sqcup 1) \times A$.
\begin{align*}
\Optic((S, S'), (A, A')) 
&= \int^{M \in \C} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
&\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
&\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
&\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
&\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
&\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S') \\
\end{align*}

\begin{remark}
Equivalently, we could let $\M$ be the category of pointed objects of $\C$ and pointed maps between them.
\end{remark}

\todo{Is it worth working the laws out?}

@@

To define setters we must recall the notion of a strong functor.
